---
title: "asthmaBiomarkers_discoveryDataProcessing_rnaSeq"
author: "Amrit Singh; Alice Yue"
date: '2017-08-07; 2018-06-14'
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
---

Logistics

```{r, echo=F, results='hide', message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=F, warning=F, message=F)

WhereAmI = "~/projects/asthma/data/RNAseq_nanostring/"
## input:
## output:
## aya43@sfu.ca
## created 20170807
## last modified 20180614

## root directory
root = "~/projects/asthma"
setwd(root)
dir.create(paste0(root, "/result"), showWarnings=F)
result_dir = paste0(root, "/result/RNAseq"); dir.create(result_dir, showWarnings=F)

## input directory
data_dir = "data/RNAseq_nanostring"

## output directory

## libraries
source("code/_func.R")
libr("dplyr")
libr("ggplot2")
libr("pracma")
libr("stringr")

```

## Compile demographics dataset

```{r}
meta_file0 = read.csv(paste0(WhereAmI, "data/allsitesDemo.csv")) %>% 
  filter(Cohort %in% c("Discovery", "Validation")) #306 --> 280 rows

## conceal subject names
meta_file0$concealed_uniqueID = paste("uniqueID",
                                       as.numeric(factor(as.character(meta_file0$NAME))),
                                       as.numeric(factor(as.character(meta_file0$AIC_YMD))), sep="_") #ID based on name + date
meta_file0$concealed_ID = paste("ID", as.numeric(factor(as.character(meta_file0$NAME))), sep="_")
```

Calculate EAR (early asthmatic response) and LAR (late asthmatic response) and compare to response column

```{r}
# scale columns
dropFEV1 = as.data.frame(t(100*scale(t(
  meta_file0[, c("BLFEV","F10L","F20L","F30L","F45L","F60L","F90L","F120L",
                 "F180L","F240L","F300L","F360L","F420L")]), 
  center=meta_file0$BLFEV, scale=meta_file0$BLFEV)))

# EAR and LAR are maximum drops (min for negative) 
# at the two groups of time intervals; rounded
ear = suppressWarnings( apply(dropFEV1, 1, function(x) {min(x[1:8], na.rm=T)}) )
ear[ear == "Inf"] = NA
lar = suppressWarnings( apply(dropFEV1, 1, function(x) {min(x[9:13], na.rm=T)}) )
lar[lar == "Inf"] = NA
meta_file0$EAR = round(ear, 1)
meta_file0$LAR = round(lar, 1)

time = gsub("L", "", substring(colnames(dropFEV1), 2))
# time = vapply(str_split(colnames(dropFEV1), "_"),'[',3, FUN.VALUE=character(1)) %>% as.numeric
time[1] = 0;
time = as.numeric(time)/60

# AUR trapz: area of a function for each patient, all and LAS time vs drops
meta_file0$AUC = apply(dropFEV1, 1, function(y) trapz(x=time, y=y) )
meta_file0$AUC_LAR = apply(dropFEV1, 1, function(y) trapz(x=time[9:13], y=y[9:13]) )

#AIS: PC20
ais = meta_file0[meta_file0$Time == "Pre", "PC20"] / meta_file0[meta_file0$Time == "Post", "PC20"]
names(ais) = meta_file0$concealed_uniqueID[meta_file0$Time == "Pre"]
meta_file0$AIS = rep(ais, each = 2)

```

Relationship between LAR and AUC

```{r}
 #LAR as a resuls of aAUC + bSubject
summary(lm(LAR ~ AUC + concealed_ID, data = meta_file0))
plot(LAR ~ AUC, data = meta_file0, pch = 19, col = 2, 
     ylab = "Max drop during LAR (3-7h) per subject", 
     xlab = "Area above the FEV1 time vs drop profile curve",
     main = "subjects (red = + subject effect)")
points(LAR ~ AUC, data = meta_file0, pch = 21, col = 1)
abline(lm(LAR ~ AUC, data = meta_file0))
text(x = 20, y = -30, labels = paste0("Pearson r = ", round(cor(meta_file0$LAR, meta_file0$AUC, use = "pairwise.complete.obs"), 2)))

## AUC of the LAR
summary(lm(LAR ~ AUC_LAR + concealed_ID, data = meta_file0))
plot(LAR ~ AUC_LAR, data = meta_file0, pch = 19, col = 2, ylab = "Max drop during LAR (3-7h)", 
     xlab = "AUC (3-7h)")
points(LAR ~ AUC_LAR, data = meta_file0, pch = 21, col = 1)
abline(lm(LAR ~ AUC_LAR, data = meta_file0))
text(x = 20, y = -30, labels = paste0("Pearson r = ", round(cor(meta_file0$LAR, meta_file0$AUC_LAR, use = "pairwise.complete.obs"), 2)))
text(x = 20, y = -40, labels = "p < 2.2e-16")

```

## Relationship between LAR and AIS

```{r}
summary(lm(AIS ~ LAR+concealed_ID, data = meta_file0))
plot(AIS ~ LAR, data = meta_file0, pch = 19, col = "green", 
     xlab = "Max drop during LAR (3-7h)", ylab="AIS (Allergen Induced Shift)")
points(AIS ~ LAR, data = meta_file0, pch = 21, col = "black")
abline(lm(AIS ~ LAR, data = meta_file0))
text(x = 5, y = 10, labels = paste0("Pearson r = ", round(cor(meta_file0$AIS, meta_file0$LAR, use = "pairwise.complete.obs"), 2)))
text(x = 5, y = 8, labels = "p = 0.60")



```

## add response variable

```{r}
meta_file0$calculated_Response = rep(NA, nrow(meta_file0))
meta_file0$calculated_Response[meta_file0$LAR > -15] = "ER"
meta_file0$calculated_Response[meta_file0$LAR <= -15] = "DR"
meta_file0$calculated_Response[as.character(meta_file0$Mac_Response) == "Control"] = "Control"

## remove samples with missing data
meta_file = meta_file0[!is.na(meta_file0$calculated_Response), ]
rownames(meta_file) = paste(meta_file$concealed_uniqueID, meta_file$Time, sep=".")
saveRDS(meta_file, paste0(WhereAmI,"/data/allsitesDemo_clean.rds"))
```

# RNA-Seq data

```{r}
WhereAmI2 =  "~/projects/asthma/data/RNAseq_nanostring/"
#################################################
#
# 1) UCSC genes
#
#################################################
fileNames = list.files(paste0(WhereAmI2, "data/discovery/rnaseq/geneCounts"), full.names=T)
gene.eset = read.delim(fileNames[1])
ucscGeneCounts0 = matrix(NA, nrow = nrow(gene.eset), ncol = 96)
rownames(ucscGeneCounts0) = as.character(gene.eset$transcript_id.s.)
colnames(ucscGeneCounts0) = list.files(paste0(WhereAmI2, "data/discovery/rnaseq/geneCounts"))
for(i in 1 : length(fileNames)){
  ucscGeneCounts0[, i] = read.delim(fileNames[i])[, "expected_count"]
}

#################################################
#
# 2) UCSC gene-isoforms
#
#################################################
fileNames = list.files(paste0(WhereAmI2, "data/discovery/rnaseq/geneIsoCounts"), full.names=T)
geneIso.eset = read.delim(fileNames[1])
ucscGeneIsoCounts0 = matrix(NA, nrow = nrow(geneIso.eset), ncol = 96)
rownames(ucscGeneIsoCounts0) = as.character(geneIso.eset$transcript)
colnames(ucscGeneIsoCounts0) = list.files(paste0(WhereAmI2, "data/discovery/rnaseq/geneIsoCounts"))
for(i in 1 : length(fileNames)){
  ucscGeneIsoCounts0[, i] = read.delim(fileNames[i])[, "expected_count"]
}

#################################################
#
# 3) Ensembl dataset
#
#################################################
starEnsembl = read.delim(paste0(WhereAmI2, "data/discovery/rnaseq/ensembl/asthma.GRCh38_ERCC.gencode21.trimmedCounts.txt"), header=F)
sampleNames = as.character(as.matrix(starEnsembl[2,-c(1:6)]))
dim(starEnsembl)   # 60249  x 102
starEnsembl.dat = starEnsembl[-c(1,2),]
colnames(starEnsembl.dat) = c(as.character(as.matrix(starEnsembl[2,c(1:6)])), gsub(".*(HI.*Index_[0-9]{1,2}.[0-9]{1,2}).*", "\\1", sampleNames))
rownames(starEnsembl.dat) = starEnsembl.dat$Geneid
starEnsemblExp00 = starEnsembl.dat[, -c(1:6)] ## 60247 (60155+92ercc) x 96

starEnsemblExp0 = matrix(0, nrow = nrow(starEnsemblExp00), ncol = ncol(starEnsemblExp00))
for(i in 1:ncol(starEnsemblExp0)){starEnsemblExp0[, i] = as.numeric(as.character(starEnsemblExp00[, i]))}
rownames(starEnsemblExp0) = rownames(starEnsemblExp00)
colnames(starEnsemblExp0) = colnames(starEnsemblExp00)
#################################################
#
# 4) Trinity dataset
#
#################################################
## contigs-isoforms
fileNames = list.files(paste0(WhereAmI2, "data/discovery/rnaseq/trinity"), full.names=T)
geneIso.eset = read.delim(fileNames[1])
trinityGeneIsoCounts0 = matrix(NA, nrow = nrow(geneIso.eset), ncol = 96)
rownames(trinityGeneIsoCounts0) = as.character(geneIso.eset$transcript)
colnames(trinityGeneIsoCounts0) = list.files(paste0(WhereAmI2, "data/discovery/rnaseq/trinity"))
for(i in 1 : length(fileNames)){
  trinityGeneIsoCounts0[, i] = read.delim(fileNames[i])[, "expected_count"]
}

## dimension of datasets
dim(ucscGeneCounts0);      #42,465 x 96
dim(ucscGeneIsoCounts0);   #89,357 x 96
dim(starEnsemblExp0);      #60,247 x 96
dim(trinityGeneIsoCounts0); # 258,403 x 96

## change dataset colnames (sample names)
colnames(ucscGeneCounts0) = unlist(gsub(".genes.results", "", gsub("rsem.UCSC_", "", colnames(ucscGeneCounts0))))
colnames(ucscGeneIsoCounts0) = unlist(gsub(".isoforms.results", "", gsub("rsem.UCSC_", "", colnames(ucscGeneCounts0))))
colnames(starEnsemblExp0) = unlist(lapply(strsplit(colnames(starEnsemblExp0), split = "_"), function(i) i[2]))
colnames(trinityGeneIsoCounts0) = unlist(gsub(".isoforms.results", "", gsub("rsem.UCSC_", "", colnames(ucscGeneCounts0))))

## Map HiSeqID to alias
mapHiSeqToID0 = read.csv(paste0(WhereAmI2, "data/demo/rnaseqHiSeqInfo.csv"))
rownames(mapHiSeqToID0) = unlist(lapply(strsplit(as.character(mapHiSeqToID0$Filename.Prefix), split = "_"), function(i) i[2]))

mapHiSeqToID = mapHiSeqToID0[order(mapHiSeqToID0$Alias), ]

## import clinical data
#meta_file = readRDS(paste0(WhereAmI, "data/demo/allsitesDemo.rds"))
#rownames(meta_file) = paste(meta_file$concealedID, meta_file$Time, sep=".")


meta_fileRNASeq00 = meta_file[meta_file$rnaseqID_quebec %in% mapHiSeqToID$Alias, ]
meta_fileRNASeq0 = meta_fileRNASeq00[order(meta_fileRNASeq00$rnaseqID_quebec), ]
all(mapHiSeqToID$Alias == meta_fileRNASeq0$rnaseqID_quebec)
meta_fileRNASeq = cbind(meta_fileRNASeq0, mapHiSeqToID)
meta_fileRNASeq$ID = unlist(lapply(strsplit(as.character(meta_fileRNASeq$Filename.Prefix), "_"), function(i){i[2]}))

## re-order samples in rna-seq datasets
ucscGeneCounts = ucscGeneCounts0[, meta_fileRNASeq$ID]
ucscGeneIsoCounts = ucscGeneIsoCounts0[, meta_fileRNASeq$ID]
starEnsemblExp = starEnsemblExp0[, meta_fileRNASeq$ID]
trinityGeneIsoCounts = trinityGeneIsoCounts0[, meta_fileRNASeq$ID]
all(meta_fileRNASeq$ID2 == colnames(ucscGeneCounts))
all(meta_fileRNASeq$ID2 == colnames(ucscGeneIsoCounts))
all(meta_fileRNASeq$ID2 == colnames(starEnsemblExp))
all(meta_fileRNASeq$ID2 == colnames(trinityGeneIsoCounts))
colnames(ucscGeneCounts) = rownames(meta_fileRNASeq)
colnames(ucscGeneIsoCounts) = rownames(meta_fileRNASeq)
colnames(starEnsemblExp) = rownames(meta_fileRNASeq)
colnames(trinityGeneIsoCounts) = rownames(meta_fileRNASeq)
```

# get RNA-Seq data ready for analysis

```{r}
source(paste0(WhereAmI2, "code/discovery/functions.R"))

## describe demographics
meta_fileRNASeq2 = meta_file[meta_file$rnaseq_coreSet_biomarkerAnalysis == "Y", ]
meta_fileRNASeq2$Time = relevel(meta_fileRNASeq2$Time, ref = "Pre")

```

## Descriptive statistics (at pre-challenge)

```{r}
meta_fileRNASeq2Pre = subset(meta_fileRNASeq2, Time == "Pre")
## Pre-challenge
variables = c("Wt..Kg.", "HT.cm.", "AGE", "BLFEV", "PRFEV", "Leukocyte_Counts.x10.9.",
              "Neu_percent", "lym_percent", "mono_percent", "eos_percent",
              "baso_percent", "EAR", "LAR", "PC20", "AIS")
summaryStat = amritr::descriptiveStat(meta_fileRNASeq2Pre, groups = "calculated_Response", variables, paired=F, pairing = NULL)
as.data.frame(summaryStat$meanSD)
as.data.frame(summaryStat$pval)

## Categorical variables
table(meta_fileRNASeq2Pre$calculated_Response, droplevels(meta_fileRNASeq2Pre$SEX))
table(meta_fileRNASeq2Pre$calculated_Response, droplevels(meta_fileRNASeq2Pre$Allergen_cleanLabel))
```

## Identity lower limit of detection; filter transcripts

```{r}
all(rownames(meta_fileRNASeq) == colnames(ucscGeneCounts))
all(rownames(meta_fileRNASeq) == colnames(ucscGeneIsoCounts))
all(rownames(meta_fileRNASeq) == colnames(starEnsemblExp))
all(rownames(meta_fileRNASeq) == colnames(trinityGeneIsoCounts))
ucscGeneCounts2 = ucscGeneCounts[, rownames(meta_fileRNASeq2)]
ucscGeneIsoCounts2 = ucscGeneIsoCounts[, rownames(meta_fileRNASeq2)]
starEnsemblExp2 = starEnsemblExp[, rownames(meta_fileRNASeq2)]
trinityGeneIsoCounts2 = trinityGeneIsoCounts[, rownames(meta_fileRNASeq2)]

## normalize data according to log2 cpm
ucscGeneCountsNorm = normalizelibSum(ucscGeneCounts2)
ucscGeneIsoCountsNorm = normalizelibSum(ucscGeneIsoCounts2)
starEnsemblExpNorm = normalizelibSum(as.matrix(starEnsemblExp2))
trinityGeneIsoCountsNorm = normalizelibSum(trinityGeneIsoCounts2)

## determine filtering cut-off based on linearity QC
ercc = starEnsemblExpNorm[60156:60247, ]
erccDat = data.frame(x = rep(as.numeric(ercc), ncol(ercc)))
erccConc0 = read.delim(paste0(WhereAmI2, "data/discovery/rnaseq/ERCC_Controls_Analysis.txt"), row.names =2)
erccConc = erccConc0[rownames(ercc), ]
erccDat2 = as.data.frame(ercc)
erccDat2$Conc = erccConc$concentration.in.Mix.1..attomoles.ul.
#pdf(paste0(WhereAmI, "figs/2-92ercc_discoveryCohort.pdf"), width = 4, height = 4)
erccDat2 %>% gather(Subj, Counts, -Conc) %>% 
  ggplot(aes(x = Conc, y = Counts)) + geom_point() +
  scale_x_log10() +
  geom_segment(aes(x = 0, xend = 14, 
                   y = 3, yend = 3), lty = 2) +
  geom_segment(aes(x = 14, xend = 14, 
                   y = -5, yend = 3), lty = 2) + 
  customTheme(sizeStripFont=10, xAngle=0, hjust=0.5, vjust=0.5, 
              xSize=10, ySize=10, xAxisSize=10, yAxisSize=10) +
  xlab("Concentration (Attomoles/uL)") + 
  ylab(expression("log"[2]~"cpm")) +
  annotate("text", x = 0.5, y = , 3.5, label = "cutoff = 3")
#dev.off()

```

# Final RNA-Seq datasets

```{r}
## remove features below 3 log2 cpm
genDats0 = list(ucscGeneCounts=ucscGeneCountsNorm,
                ucscGeneIsoCounts=ucscGeneIsoCountsNorm,
                starEnsemblExp=starEnsemblExpNorm,
                trinityGeneIsoCounts=trinityGeneIsoCountsNorm)
genDats = lapply(genDats0, function(x){
  y = x
  y[y < 3] = NA
  na.omit(y)
})
lapply(genDats, dim)

library(rafalib)
## sample histograms
plotSampleHist(genDats$ucscGeneCounts, "UCSC genes", xlim=c(-5, 20), ylim=c(0, 800))
plotSampleHist(genDats$ucscGeneIsoCounts, "UCSC gene-isoforms", xlim=c(-5, 20), ylim=c(0, 800))
plotSampleHist(genDats$starEnsemblExp, "Ensembl", xlim=c(-5, 20), ylim=c(0, 700))
plotSampleHist(genDats$trinityGeneIsoCounts, "Trinity", xlim=c(-5, 20), ylim=c(0, 700))

```

## save datasets to file

```{r}
rnaseq_meta_file = meta_fileRNASeq2
rnaseq_datasets = genDats
lapply(rnaseq_datasets, function(i){
  all(colnames(i) == rownames(rnaseq_meta_file))
})

## add RNA-Seq HiSeq info and demographics
mapHiSeqToID$rnaseqID_quebec = mapHiSeqToID$Alias
rnaseq_meta_file2 = inner_join(x = rnaseq_meta_file, y = mapHiSeqToID, by = "rnaseqID_quebec")
all(rnaseq_meta_file2$rnaseqID_quebec == rnaseq_meta_file$rnaseqID_quebec)
rownames(rnaseq_meta_file2) = rownames(rnaseq_meta_file)
rnaseq_meta_file = rnaseq_meta_file2

save(rnaseq_meta_file=rnaseq_meta_file, rnaseq_datasets=rnaseq_datasets,
     file = paste0(WhereAmI, "data/rnaseqDatasets.RDATA"))

```

